---
title: "Project Causal Inference"
author: "Group 14 (Ariana, Jiraiyu, Ravi, Yashish)"
date: "4/17/2022"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(MatchIt)
```

## 1 Executive Summary

AIDS infection in Africa is on the rise. One of the strategies to
prevent this is to make sure the people getting tested are getting their
results. In this document we evaluate an experiment to explore if giving
incentives to people gets them to find out the results. We have done so
using Propensity Score Matching technique and OLS regression. The
results indicate that giving incentive indeed pushes people to find
their results. Also, the distance to the VCT center is another factor
which impacts decision to find the test result.

## 2 Background

Over the past two decades, the HIV/AIDS epidemic has afflicted millions
of individuals in Africa. In the absence of significantly expanded
prevention and treatment programs, the epidemic is expected to worsen in
many other parts of the world. One intervention often suggested to
alleviate the spread of the disease is HIV testing, and some have gone
so far as to say that voluntary counseling and testing (VCT) is the
"missing weapon in the battle against AIDS". Under the assumption that
HIV testing is an effective prevention strategy, many international
organizations and governments have called for increased investments in
counseling and testing, requiring large amounts of monetary and human
resources. For example, in South Africa, government expenditures on
counseling and testing increased from \$2.4 million in 2000 to \$17.3
million in 2004, and in Mozambique, 55 percent of all HIV/AIDS program
expenditures in 2000 were for HIV counseling and testing.

Underlying the emphasis on HIV testing for prevention---and the
large-scale expenditures on testing---are two rarely challenged
assumptions. First, many believe that knowledge of HIV status has
positive effects on sexual behavior that prevent the spread of the
disease. In particular, it is assumed that those diagnosed HIV-negative
will protect themselves from infection and those diagnosed HIV-positive
will take precautions to protect others. Second, many believe that it is
difficult to get people to learn their HIV status, due primarily to
psychological or social barriers, thus justifying expenditures on
destigmatization and advertising campaigns. In this project, we evaluate
a field experiment in rural Malawi designed to address these assumptions
where in respondents were offered free door-to-door HIV test and were
randomly assigned vouchers (\$0-\$3) redeemable upon obtaining the
results at a nearby VCT center.

## 3 Data Overview

1.  Overall data had 4820 observations and 9 features which are as
    follows:

    a.  Gender (male)

    b.  Unique Village Id (villnum)

    c.  Got/Not got results (got)

    d.  Received/Not Received incentives (any)

    e.  Total Incentives (tinc)

    f.  Distance to VCT center in Kilometers (distvct)

    g.  Married or Not (mar)

    h.  Age (age)

    i.  HIV Test Result (hiv2004)

2.  Summary statistics (appendix 9.2):

    Number of respondents: 4820

    Number of Males: 2260

    Age range (Males): 13 to 84

    Number of Females: 2560

    Age Range (Females): 11 to 81

    Survey Response Rates: 67%

    HIV Test Acceptance: 90.49%

    Learned Result: 69.66%

    HIV Prevalence: 6%

    Incentives Received: 2016

    No Incentives: 878 Average Age: 33

## 4 The Experiment

Free door-to-door tests were conducted to collect HIV samples. Men and
women were randomly selected from approximately 140 villages located in
the Northern, Central, and Southern regions of Malawi from,
approximately, 25 percent of all households in each village. Between May
and August of 2004, nurses from outside each area offered respondents
free tests in their homes for HIV (samples were taken through oral
swabs). Vouchers ranging from 0 to \$3 were drawn randomly from the bag
with an average amount of \$1.01. For each subject, VCT center, to
receive the results, were randomly selected.

## 5 Analysis

#### 5.1 Randomization Test

Randomization in an experiment refers to a random assignment of
participants to the treatment in an experiment. It makes sure that the
groups made for conducting an experiment are as similar as possible to
each other so that the results come out as accurate as possible.
Randomization prevents biases and eliminates any possibility of an
unobserved confound and possibility of simultaneity. To check if the
subjects selected for the experiments (appendix 9.3) were completely
random, we ran 't.test' for each of the features between control and
treatment. From the results of the experiment we could say that, except
with respect to age (p-value = 0.005421), in all the features the
subjects chosen were completely random. But since the age was not
randomized in the data, we decided to match subjects in control and
treatment using propensity score matching.

#### 5.2 Propensity Score Matching (PSM)

Propensity score matching is a statistical matching technique that
attempts to estimate the effect of a treatment by accounting for the
co-variates that predict receiving the treatment. PSM attempts to reduce
the bias due to confounding variables that could be found in an estimate
of the treatment effect obtained from simply comparing outcomes among
units that received the treatment versus those that did not. In PSM, we
find matched pairs that are as similar on observed variables, so that
the treatment can be assumed to be random. If no match is found, throw
away the data. In PSM we model treatment as a function of observed
variables using logit or probit. We then get predicted probabilities of
treatment from the model and for each treated observation, find control
observation with closest propensity score.

To simulate randomization and balance the number of subjects in
treatment and control, PSM matches each individual in control and
treatment based on given features (age, distance to VCT center, gender
and if married or not in this case) using a propensity score (appendix
9.4). After propensity score matching we re-tested the randomization for
each figure and there are no columns which fail randomization.

#### 5.3 Power of the Experiment

We needed to check whether our experiment was under-powered i.e. having
a sample that is too small. If the sample lacks power, it means it
cannot reliably detect effects and the chance of discovering effects
that aren't there goes up. This could be because random noise in the
data leads to more chance differences in small samples. To report a
difference of 0.45 between control and treatment. with 95% level of
significance and 80% power, we need at least 79 respondents in both
treatment and control (appendix 9.5). In the matched data, we have 400+
rows of data both the groups. Thus the experiment is sufficiently
powerful.

#### 5.4 OLS Regressions

Initially regressed age, marital status, gender, distance to VCT center,
and incentives over if the subject proceeded to get the result. Then,
regressed incentives over if the subject proceeded to get the result.
Next, regressed distance to VCT center over if the subject proceeded to
get the result. Lastly, regressed the incentive the subject got over if
the subject proceeded to get the result.

We found that:

1\. Any amount of incentive increases the odds of subjects finding out
the results by almost 7 times.

2\. Increasing incentive amount by \$1 increases the odds of subject
finding out the results by about 1.01 times.

3\. Decreasing the distance to VCT center by 1 km, increases the odds of
subject finding out the results by about 1.2 times.

Please refer to appendix 9.7 for all the results.

## 6 Assumptions

We have had below assumptions which have been tested in the Analysis
section:

1.  Randomization of the features in the experiment subjects
2.  The experiments has enough power to provide the significant results.

## 7 Threats and Limitations to the Experiment

1.  Omitted variable bias : There could be other factors, such as
    behavior of VCT center staff due to which a subject might be wary of
    getting the results
2.  Selection Bias : The design of this experiment avoids selection bias
    because it randomized: Individual incentives to learn HIV status
    Location of VCT centers where HIV results were available
3.  Measurement error: There is no measurement error as experiment also
    measured actual post-test attendance at centers to obtain results
4.  Simultaneity: There can be no source of simultaneity

## 8 Conclusions and Recommendations

From the above experiments we have established causality between getting
the getting incentives and if the subject proceeds to get the results.
Since incentives increase the odds of a subject finding out their test
results, long term programs should be developed to provide incentives to
all patients who are getting test for HIV. And, since decreasing the
distance to VCT increases the odds of subject finding out the results,
infrastructure should be set up to increase the number of centers. This
would make the centers more accessible for the subject to get the
results. These two combined with each other could be an effective HIV
prevention strategy in rural Malawi.

## 9 Appendix

#### 9.1 Load data

```{r, echo=FALSE}
all_data = read.csv("HIV_Data (1).csv")

all_data = all_data %>% select(villnum, got, distvct, hiv2004, tinc, any, age, male, mar)

final_data = na.omit(all_data)
```

Column 'any' tells us if the subject is treatment or control.

#### 9.2 Summary statistics of the data

```{r}
### Count of all respondents
summary_data = all_data %>% 
    mutate(villnum = as.factor(villnum),
           got = as.factor(got),
           any = as.factor(any),
           male = as.factor(male),
           mar = as.factor(mar))
summary(summary_data)

```

#### 9.3 Randomization check

```{r}
treatment_data = final_data %>%
                 filter(any == 1)

control_data = final_data %>%
                 filter(any == 0)

print(t.test(treatment_data$age, control_data$age))
print(t.test(treatment_data$distvct, control_data$distvct))
print(t.test(treatment_data$male, control_data$male))
print(t.test(treatment_data$mar, control_data$mar))
```

#### 9.4 Propensity score matching

```{r}
match_output <- matchit(any ~ age + distvct + male + mar, 
                        data = final_data, 
                        method = 'nearest', 
                        distance = "logit", 
                        caliper = 0.001, 
                        replace = FALSE, 
                        ratio = 2)
print(summary(match_output))
data_match = match.data(match_output)

```

#### 9.5 Randomization re-check for matched data

```{r}
treatment_data = data_match %>%
                 filter(any == 1)

control_data = data_match %>%
                 filter(any == 0)

print(t.test(treatment_data$age, control_data$age))
print(t.test(treatment_data$distvct, control_data$distvct))
print(t.test(treatment_data$male, control_data$male))
print(t.test(treatment_data$mar, control_data$mar))
```

#### 9.6 Power Test

```{r}
powerTestResults = power.t.test(type = 'two.sample', 
                                power = .8, 
                                sig = 0.05, 
                                delta = .45, 
                                sd = 1)
powerTestResults
```

#### 9.7 OLS Regressions

```{r}
model = glm(got ~ distvct + any + age + male + mar,
           data = data_match,
           family = "binomial")
summary(model)

model = glm(got ~ any,
           data = data_match,
           family = "binomial")
summary(model)

model = glm(got ~ any + distvct,
           data = data_match,
           family = "binomial")
summary(model)

model = glm(got ~ tinc + distvct,
           data = data_match,
           family = "binomial")
summary(model)

model = glm(got ~ tinc + distvct,
           data = data_match,
           family = "binomial")
summary(model)
```

## 10 References

1.  Data:
    <https://github.com/NickCH-K/causaldata/blob/main/Python/causaldata/thornton_hiv/data.py>
2.  Actual experiment :
    <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3115776/>
